---
title: "Exam 1"
author: "STAT 251 Section 01"
output:
 pdf_document:
   keep_tex: true
   number_sections: true
header-includes:
  \usepackage{float}
#bibliography: home2bibs.bib
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=10, fig.pos = "H")
#knitr::opts_chunk$set(list(echo = FALSE, eval = FALSE))
library(kableExtra)
library(knitr)
library(latex2exp)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(plyr)
mainpath = 'C:/Users/Bruin/Desktop/GS Academia/TA/Teaching STAT 251/FALL/'
wmpath = 'C:/Users/Bruin/Desktop/GS Academia/SEM 2/STAT 519/PROJ 1/watermelon/simulated datasets/ wawamelon/WaterMelonMaster.csv'
whiskpath = 'C:/Users/Bruin/Desktop/GS Academia/SEM 2/STAT 519/PROJ 1/Whiskers/SimSealData/whiskerMaster.csv'
```







\begin{table}
\centering
\begin{tabular}{cccc}

Student Name: & $\_\_\_\_\_\_\_\_\_\_\_\_\_\_$ & Last Four of Vandal Number:  & $\_\_$ $\_\_$ $\_\_$ $\_\_$ \\
&&& \\
Test Version: & \textbf{A} && \\

\end{tabular}
\end{table}


\newpage


\begin{enumerate}
\item[ 2pts \bf 3.)]{Multiple choice: A plot which shows distinct values of a variable on the $x$-axis and the frequency or relative frequency on the $y$-axis is called a:}
\item[(a)]{Dot plot}
\item[(b)]{Pie chart}
\item[(c)]{Box and Whisker Plot}
\item[(d)]{Histogram}
\end{enumerate}


\newpage
\begin{enumerate}
\item[2pts \bf 4.)]{Multiple choice: A preliminary exploration and summary of the data}
\item[(a)]{descriptive statistics}
\item[(b)]{sampling design}
\item[(c)]{interquartile range}
\item[(d)]{box and whisker plot}
\end{enumerate}


\begin{enumerate} 
  \item[2pts \bf 5.)]{True or False: Inferential statistics can be applied to both samples and populations}
  \item[] Answer:$\_\_\_\_\_\_$
\end{enumerate}



\begin{enumerate} 
  \item[4pts \bf 6.)]{Write in the letter(s) of the words or phrases that best complete the following sentence: "The mean is the \_\_\_\_\_ of a distribution while the median is the \_\_\_\_\_ of a distribution."}
  \item[(a)] middle value
  \item[(b)] frequency
  \item[(c)] measure of spread
  \item[(d)] center of gravity
\end{enumerate}


\begin{enumerate} 
  \item[4pts \bf 7.)]{Write in the letter(s) of the words or phrases that best complete the following sentence: "The range is a \_\_\_\_\_ of a distribution. It is very \_\_\_\_\_ to outliers"}
  \item[(a)] measure of spread
  \item[(b)] susceptible
  \item[(c)] measure of location
  \item[(d)] resistant
\end{enumerate}











\newpage
\begin{enumerate}
\item[14pts \bf 8.)] Consider the following data from a survey conducted on sample of 20 college students in the state of Georgia. The survey, conducted by the State, was investigating several variables to investigate how student behavior affects academic performance. One of the survey questions required students to self report the amount of time (in whole hours) they spend studying each day. The following frequency table gives the distribution of the number of hours spent studying. Fill in the table (5 pts) and answer questions (a)-(c):
\end{enumerate}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ix = c(25,39,53,57,22,43,36,10,20,41,58,3,1,13,15,49,9,4,16,29)
gadata = read.csv(paste0(mainpath, '/Week_2/georgia_student_survey.csv'))[ix,]
gadata$Studytime = round_any(gadata$Studytime, 1)

st = sort(unique(gadata$Studytime))
st_F = summary(as.factor(gadata$Studytime))
st_RF = st_F/sum(st_F)
st_CRF = cumsum(st_RF)


FT = cbind.data.frame(`Study Time (Hrs)` = st, 
                      Frequency = c(st_F[1],"", st_F[3:10]),
                      `Relative frequency` = c(st_RF[1:2], "", "", "", st_RF[6:10]),
                      `Cumulative RF` = c(st_CRF[1:3], rep("", 7)))

kable(FT, format = 'latex', digits = 2, row.names = F)
```


\begin{itemize}
\item[2pts (a)]{What kind of variable is ``Study Time (Hrs)"?}
\item[]
\item[]
\item[5pts (b)]{Using the frequency table above, compute the mean amount of time a student spends studying in this sample: show your calculations}
\item[]
\item[]
\item[]
\item[]
\item[]
\item[]
\item[]
\item[]
\item[2pts (c)]{What proportion of students study at most 5 hours a day?}
\end{itemize}

\newpage
\begin{enumerate}
\item[5pts \bf 9.)]  construct a histogram 
\item[]
\item[]
\item[]
\item[]
\item[]
\item[]
\item[]
\end{enumerate}




\begin{enumerate}
\item[10pts \bf 10.)]{consider the following sample of 3 observations of a quantitative variable $X$ 
\[ X = \{-4.4, 1.5, -0.2 \}\]}
\item[a.] Show how to compute the sample variance
\item[]
\item[]
\item[]
\item[]
\item[b.] Assuming the variance of $X$ above $s^2 \approx 9.22$, what is the standard deviation $s$? 
\item[]
\item[]
\item[]
\end{enumerate}





\begin{enumerate}
\item[\bf 13.)] match boxplot to correct histogram distribution
\end{enumerate}






\newpage
\begin{enumerate}
\item[18pts \bf 14.)] \textbf{Table 1.} gives the distribution of a quantitative variable $x$. Use this table to answer parts (a) and (b)
\end{enumerate}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
v = c(5,6,7,8,9,10,11,12)
fv = c(8, 5, 2, 15, 7, 8, 5, 10)
sv = sum(fv)
rfv = fv/sv
crfv = cumsum(rfv)
tab = cbind.data.frame(X = v, `Frequency(X)` = fv, `Relative Frequency(x)` = rfv, `Cumulative Relative Fequency` = crfv)

kable(tab, format = 'latex', booktabs = T, caption = '', digits = 3)
rv = c(rep(5,8), rep(6,5), rep(7,2), rep(8,15), rep(9,7), rep(10,8), rep(11,5), rep(12,10))

```


\begin{enumerate}
\item[9pts (a)]{Plot the cumulative distribution and use this plot to find the $25th$, $50th$, and $75th$ percentiles}
\end{enumerate}

```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}

df = cbind.data.frame(x = rv, y = runif(60))
ggplot(data = df, aes(x =x, y = y))+
  geom_blank()+
  theme_hc()+
  scale_x_continuous(breaks = seq(5, 12, 1))+
  scale_y_continuous(breaks = seq(0, 1, 0.1))+
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 15),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 15),
        panel.grid.major.y=element_line(colour="black"))+
  ylab('')+
  xlab('')
  
  
```

$25th = \_\_\_\_$  $50th = \_\_\_\_$  $75th = \_\_\_\_$ 

\newpage
\begin{enumerate}
\item[9pts (b)] Below is a box and whisker plot for the frequency table in \textbf{table 1}. Using the relevant statistics, label the values of the minimum, maximum, Q1, Q3, and IQR [hint you can use your calculations from part (a)]. 
\end{enumerate}

```{r, echo=FALSE, message=FALSE, warning=FALSE}

Q1 = 7
med = 8+0.5
Q3 = 10
IQR = 3
minv = 5+.2
maxv = 12-.2
estims = c(minv, Q1, med, Q3, maxv)


boxplot(rv, horizontal = TRUE, axes = FALSE, staplewex = 1)
text(x=estims, labels =paste(c("Min", "Q1","Median", "Q3", "Max"), "_____", sep = "="), 
     y=1.25, cex = 1.5)
text(x=8, labels =paste("IQR", "_______", sep = " = "), y=1.5, cex = 2)
```



\begin{enumerate}
\item[ 4pts \bf(bonus)]{Why is the sample variance divided by $n-1$ instead of $n$ like the sample mean? explain you answer (a mathematical demonstration can also help)}
\end{enumerate}
