---
title: "Exam 1 Review Problems"
author: "STAT 251 Section 03"
date: "2024-01-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(latex2exp)
path = 'C:/Users/Bruin/Desktop/GS Academia/TA/Teaching STAT 251/FALL/Week_1/'
path2 = 'C:/Users/Bruin/Desktop/GS Academia/TA/Teaching STAT 251/FALL/Week_2/'
```

${\bf (1)}$ For each statistic, denote whether it is resistant or not resistant to outliers:
<ol type="a">
  <li>sample mean $\bar{x}$ $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>median $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>quartiles $Q1, Q3$ $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>range $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>interquartile range (IQR) $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>variance $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
  <li>standard deviation $\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_$</li>
</ol>

${\bf (2)}$

A survey of college students in Georgia asked the following questions:
<ol type="a">

  <li>Are you male or female? (recorded as male = 0, female = 1)</li>
  
  <li>What is you height in inches?</li>
  
  <li>Do you routinely smoke cigarettes? (recorded as no = 0, yes = 1)</li>
  
  <li>How many hours do you spend studying?</li>
</ol>

The figure below shows histograms of the student responses to each of these questions in scrambled order and without scale markings. Which histogram goes with each variable? Explain your reasoning.

```{r, echo=FALSE, message=F, warning=F, fig.height=5, fig.width=8}

gss = read.csv(paste0(path2, 'georgia_student_survey.csv'))
k = ceiling(sqrt(dim(gss)[1]))

A = ggplot(data = gss, aes(x = Gender))+
  geom_histogram(color = 'black', fill = 'navy')+
  theme_classic2()+
  xlab('')+
  ylab('Frequency')+
  ggtitle('Variable: ______________')+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

B = ggplot(data = gss, aes(x = Studytime))+
  geom_histogram(color = 'black', fill = 'navy', position = 'dodge', bins = k)+
  theme_classic2()+
  xlab('')+
  ylab('Frequency')+
  ggtitle('Variable: ______________')+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


C = ggplot(data = gss, aes(x = Height))+
  geom_histogram(color = 'black', fill = 'navy', position = 'dodge', bins = k)+
  theme_classic2()+
  xlab('')+
  ylab('Frequency')+
  ggtitle('Variable: ______________')+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


D = ggplot(data = gss, aes(x = Smokecig))+
  geom_histogram(color = 'black', fill = 'navy')+
  theme_classic2()+
  xlab('')+
  ylab('Frequency')+
  ggtitle('Variable: ______________')+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


ggarrange(A, B, C, D, nrow = 2, ncol = 2)

```


<br>
<br>
<br>
<br>
${\bf (3)}$
Consider the following 10 observations of measurements made on Egyptian skulls from five epochs. Use this data to answer parts (a) - (d).

```{r, echo=FALSE, message=F, warning=F, fig.height=5, fig.width=8}

set.seed(555)
library(HSAUR2)

kable(skulls[sample(1:dim(skulls)[1], 10),], format = 'html', digits = 2, row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')


```

<ol type="a">

  <li>What type of variable is "Epoch"?</li>
  
  <li>Fill in the frequency table below for the variable "Epoch"</li>
  

```{r, echo=FALSE, message=F, warning=F, fig.height=5, fig.width=8}
set.seed(555)
x = skulls[sample(1:dim(skulls)[1], 10),]
cts = summary(x$epoch)[which(summary(x$epoch)>0)]

tbl = cbind.data.frame(Epoch = names(cts), Frequency = rep("",4))

kable(tbl, format = 'html', digits = 2, row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')


```

<li>Use the frequency table from part a.) to construct a bar plot of the variable "Epoch"</li>



```{r, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=6}
set.seed(555)
x = skulls[sample(1:dim(skulls)[1], 10),]
cts = summary(x$epoch)[which(summary(x$epoch)>0)]

tbl = cbind.data.frame(names(cts), Frequency = rep("",4))

ggplot()+geom_blank()+
  theme_bw()+
  scale_y_continuous(limits = c(0, 0.6), breaks = seq(0, 0.6, 0.1))+
  xlim(names(cts))+
  ylab('Relative Frequency')+
  xlab('Epoch')


```

<li>What proportion of the observations in the sample are from the epoch "c150AD"?</li>

</ol>


<br>
<br>
<br>
<br>
${\bf (4)}$
Again, consider the 10 observations of measurements made on Egyptian skulls from five epochs given in problem $\bf (1)$. Answer parts (a)  (c) 
<ol type="a">
  <li>Construct a histogram for the variable $nh$ (nasal height) using the following intervals: $48 < X \leq 50$, $50< X \leq 52$, $52< X\leq 54$, and $54< X\leq 56$. Start by filling in the frequency table below</li>


```{r, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=4}
set.seed(555)
x = skulls[sample(1:dim(skulls)[1], 10),]
cts = c(sum(x$nh<=50), sum(x$nh>50 & x$nh<=52), sum(x$nh>52 & x$nh<=54), sum(x$nh>54))

tbl = cbind.data.frame(Interval = c('$48 < X \\leq 50$','$50< X \\leq 52$', '$52< X \\leq 54$', '$54< X \\leq 56$'), Frequency = rep("",4))

kable(tbl, format = 'html', digits = 2, row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

ggplot()+geom_blank()+
  theme_bw()+
  scale_y_continuous(limits = c(0, 4), breaks = seq(0, 4, 1))+
  scale_x_continuous(limits = c(min(x$nh), 56), breaks = seq(min(x$nh), 56, 2))+
  ylab('Frequency')+
  xlab('Nasal Height')


```
<li>Using the histogram above, what proportion of the skulls have a nasal height less than or equal to 54 milimeters?</li>

<li>Describe the shape of the distribution above</li>
</ol>

<br>
<br>
<br>
<br>
${\bf (5)}$

Consider the following $15$ observations from an experiment on the growth of orange trees. Note that the age in years has been rounded to the nearest $1/2$
```{r, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=4}

set.seed(555)

x2 = Orange[sample(c(1:dim(Orange)[1]), 15),]

ix = sort(x2$age, index.return = T)$ix

x2 = x2[ix,]

tbl = cbind.data.frame(`Age (years)` = plyr::round_any(x2$age/365, 0.5), `Circumference (mm)` = x2$circumference)
kable(tbl, format = 'html', digits = 2, row.names = T, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')
```

<ol type="a">
  <li>What type of variable is "Age"?</li>
  <li>What type of variable is "Circumference"?</li>
  <li>Give the 5 number summary for the variable "Age"?</li>
  <li>Construct a boxplot using the five number summary above, be sure to make note of any outliers</li>
  
```{r, echo=FALSE, message=F, warning=F, fig.height=2, fig.width=5}

set.seed(555)

x2 = Orange[sample(c(1:dim(Orange)[1]), 15),]
x2$age = plyr::round_any(x2$age/365, 0.5)
ggplot()+geom_blank()+
  theme_classic2()+
  scale_x_continuous(limits = c(min(x2$age), max(x2$age)), breaks = seq(min(x2$age), max(x2$age), 0.5))+
  xlab('Orange Tree Age (years)')+
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
</ol>


<br>
<br>
<br>
<br>
${\bf (6)}$

Consider the following frequency table for the magnitude of earthquakes recorded near the island nation of Fiji. 
```{r, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=4}

set.seed(432)

magr = plyr::round_any(quakes$mag, 0.5)
xvals = unique(magr)
cts = summary(as.factor(magr))
RF = cts/sum(cts)
CRF = cumsum(RF)

which = sample(1:6, 5, replace = T)
cts2 = cts
cts2[which[1]] = ''

RF2 = RF
RF2[which[2:3]] = ''

CRF2 = CRF
CRF2[which[4:5]] = ''
tbl = rbind(cbind.data.frame(Magnitude = xvals, Frequency = cts2, `Relative Frequency` = RF2, `Cumulative Relative Frequency` = CRF2 ), c("Total", sum(cts), sum(RF), 1.0))
kable(tbl, format = 'html', digits = 2, row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')%>%
  column_spec(4, width = "18em")


```

<ol type="a">
  <li>Complete the frequency table above</li>
  <li>Confirm that the mean magnitude of an earthquake is about $4.6$</li>
  <li>Plot the cumulative distribution of earthquake magnitude</li>
```{r, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=6}
set.seed(432)

magr = plyr::round_any(quakes$mag, 0.5)
xvals = unique(magr)
ggplot()+geom_blank()+
  theme_bw()+
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1))+
  scale_x_continuous(limits = c(min(xvals), max(xvals)), breaks = seq(min(xvals), max(xvals), 0.5))+
  ylab('Cumulative Relative Frequency')+
  xlab('Earthquake Magnitude')

```
  <li>Using the plot above, what the 75th percentile of the earthquake magnitude?</li>
</ol>


