---
title: "Homework 5"
output:
  html_document: default
  pdf_document: default
always_allow_html: true
header-includes:
  \usepackage{float}
bibliography: C:/Users/Bruin/Documents/GitHub/UI_STAT251-03_Spring_2024/docs/misc/home5bibs.bib
---




```{r setup, include=FALSE}
set.seed(555)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=10, fig.pos = "H")
#knitr::opts_chunk$set(list(echo = FALSE, eval = FALSE))
library(kableExtra)
library(knitr)
library(latex2exp)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(plyr)
mainpath = 'C:/Users/Bruin/Documents/GitHub/UI_STAT251-03_Spring_2024/docs/Data/'
wmpath = 'C:/Users/Bruin/Desktop/GS Academia/MS_Stat/SEM 2/STAT 519/PROJ 1/watermelon/simulated datasets/ wawamelon/WaterMelonMaster.csv'
whiskpath = 'C:/Users/Bruin/Desktop/GS Academia/MS_Stat/SEM 2/STAT 519/PROJ 1/Whiskers/SimSealData/whiskerMaster.csv'

datapath = 'C:/Users/Bruin/Documents/GitHub/UI_STAT251-03_Spring_2024/docs/Data/'
#<span style="color: red;"></span>
```


### Definitions
Define the following terms/concepts

* a.) **Non-parametric statistics**

* b.) **Parametric statistics**

* c.) Describe the advantages and disadvantages of **Non-parametric** significance tests

* d.) **Explanatory Variable**

* e.) **Response Variable**


### Confidence intervals and two-tailed tests:

$\bf (1.)$ For parts (a) - (d) given the confidence interval for the parameter under the null hypothesis, use the point estimate to determine whether the null hypothesis $H_0$ should be **Rejected** or **Not Rejected** and denote significance level $\alpha$

a.) $H_0: \mu_0 = -2$ <br> 
  $H_A: \mu \neq \mu_0$ <br> 
  $95\%$ CI for $\mu_0 \in [-3.0, -2.4]$ <br> 
  $\bar{x} = -2.7$
  \[ \text{Decision} = \_\_\_\_\_\_\_ \\
  \alpha = \_\_\_\_\_\_ \] 

b.) $H_0: p_0 = 0.9$ <br> 
$H_A: p \neq p_0$ <br>
$99\%$ CI for $p_0 \in [0.972, 0.968]$ <br>
$\hat{p} = 0.97$ 

\[ \text{Decision} = \_\_\_\_\_\_\_ \\
  \alpha = \_\_\_\_\_\_ \]

c.) $H_0: \mu_0 = 0$ <br> 
$H_A: \mu \neq \mu_0$ <br>
$90\%$ CI for $\mu_0 \in [-3.63, 7.49]$ <br> 
$\bar{x} = 1.93$

\[ \text{Decision} = \_\_\_\_\_\_\_ \\
  \alpha = \_\_\_\_\_\_ \]

d.) $H_0: \mu_0 = 50$<br> 
$H_A: \mu \neq \mu_0$<br>
$85\%$ CI for $\mu_0 \in [93.6, 139]$<br> 
$\bar{x} = 116.3$

\[ \text{Decision} = \_\_\_\_\_\_\_ \\
  \alpha = \_\_\_\_\_\_ \]

### The sign test

<br>
<br>

$\bf (2.)$ A dermatologist is conducting an experiment to test whether a new topical skin care cream reduces the appearance facial of acne. From a sample of $20$ patients with acne the researcher gives each patient a score from $1-5$ where (5) representing the severity of their acne before the topical cream is applied. Patients are then asked to use the topical cream for $10$ days. Following the $10$ day treatment, patients are given a new acne score between $1-5$. The data for this experiment is given in the table below:



```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(555)
x = sample(c(1:5), 20, prob = c(0.1, 0.2, 0.2, 0.2, 0.3), replace = T)
tab = cbind.data.frame(`Patient Number` = c(1:20), 
                       `Score Before Treatment` = x,
                       `Score After Treatment` = sample(c(1:5), 20, prob = c(0.3, 0.3, 0.2, 0.1, 0.1), replace = T),
                       Difference = rep("", 20),
                       Sign = rep("", 20))

kable(tab, format = 'html', digits = 3,
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')
```

The dermatologist is particularly interested in whether the new product can improve the scars' appearance. They plan to test the null hypothesis $H_0: p_0 = 0.5$ using a right-tailed test with alternative hypothesis $H_A: p>0.5$ using a sign test to determine if there is enough evidence to conclude that the topical cream a positive effect on reducing facial acne at the $\alpha = 0.05$ level.

  a.) Fill in the table above by computing the difference between in acne scores before and after applying the skin care cream. Record the sign of each difference and report the total number of positive signs $s$

  b.) What does a "positive" sign indicate in this experiment? 

  c.) Compute the $p$-value for the sign test

  d.) Use the $p$-value from part (c) to determine if there is enough evidence to conclude that the topical cream product has a positive effect on reducing the appearance of acne. Interpret your decision in context. 

<br>
<br>

$\bf (3.)$ In a boreal forest, researchers conducted an experiment to study how herbivores respond to variations in food plant quality and quantity at the stand level. They fertilized young forest stands and observed herbivore use over the subsequent year. The data collected included the number of animal tracks in in the fertilized and control plots [@ball2000response]

```{r, echo=F, warning=F, message=F}
herbivore_data <- cbind.data.frame(
  Observation = 1:12,
  `Number of Tracks In Fertilized plots` = c(15, 12, 18, 14, 16, 13, 17, 14, 19, 15, 11, 16),
  `Number of Tracks In Control Plots` = c(10, 9, 11, 8, 12, 10, 11, 9, 13, 10, 8, 12),
  `Difference` = rep('',12),
  `Sign` = rep('',12)
)

kable(herbivore_data, format = 'html', digits = 2,
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```

The researchers are interested in whether or not fertilizing stands increased herbaceous activity. Conduct a sign test at the $\alpha = 0.05$ signifcance level to determine if fertilized plots have significantly more herbivore tracks than control plots



a.) Fill in the table above by computing the difference in animal tracks between the fertilized and control plots. Record the sign of each difference and report the total number of positive signs $s$

  b.) What does a "positive" sign indicate in this experiment? 

  c.) Compute the $p$-value for the sign test

  d.) Use the $p$-value from part (c) to determine if there is enough evidence to conclude that fertilized stands have more herbivore activity.
  
<br>
<br>

### Comparing two population proportions $p_1$ and $p_2$


$\bf (4.)$ The United States Supreme Court serves as the judicial branch of the U.S. government, tasked with ensuring that laws conform to the U.S. Constitution. Although traditionally operating as a relatively discreet arbiter within the American government, recent years have seen an unusual spotlight on the Supreme Court in partisan politics. This heightened attention is largely attributed by political analysts to several landmark Supreme Court decisions and ethical controversies that have surfaced. Consequently, it is believed that this increased scrutiny has adversely affected public approval of the Court across the political spectrum. The table below provides a summary of current and historical polling results for assessing SCOTUS job approval. The current approval rating is summarized from $192$ political polls conducted among the public since late 2022 [@RyanBest_2024]. The historic approval sentiment is characterized from $39$ Gallup polls conducted since early 2001. 

```{r, echo=F, warning=F, message=F}
poll_data = read.csv(file = paste0(datapath,'scotus_approval_polls.csv'))
hs_poll_data_gallup = read.csv(file = paste0(datapath,'historical_scotus_approval.csv'))

sumtab = cbind.data.frame(`Period` = c('Current Approval (2022 - 2024)', 'Past Approval (Gallup: 2001 - 2022)'),
                          `Parameter` = c('$p_1$','$p_2$'),
                          `Sample Size` = c(dim(poll_data)[1], dim(hs_poll_data_gallup)[1]),
                          `Mean SCOTUS Approval Rating` = c(mean(poll_data$yes), mean(na.omit(hs_poll_data_gallup$Approve))))

kable(sumtab, format = 'html', digits = 2,
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```


Conduct a two-sample test at the $\alpha = 0.05$ significance level to determine whether the current approval rating is significantly lower than the historical average - be sure to address all five steps of the hypothesis test.

<br>
<br>

$\bf (5.)$ A study published in the <i>Journal of Paediatrics and Child Health</i> was interested in the number of reports of child abuse in Sergipe, Brazil, before and after COVID-19 started. The researchers hypothesized that children in abusive homes might be in more danger because they weren’t going to school and their families were under more stress [@martins2020decrease]. A summary of the study is reported below:

```{r, echo=F, warning=F, message=F}
df = cbind.data.frame(Period = c('Before COVID-19 (2019)', 'After COVID-19 (2020)'),
                      `Parameter` = c('$p_1$','$p_2$'),
                      `Total Registries` = c(70,53),
                      `Registries for Childer under 12 Years of Age` = c(18, 16),
                      `Relative Proportion` = c(round(18/70,3), round(16/53,3)))

kable(df, format = 'html', digits = 2,
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')


```

Conduct a two-sample test at the $\alpha = 0.01$ significance level to determine whether the rate of reported child abuse for children under the age of $12$ was significantly different after the start of the COVID-19 pandemic - be sure to address all five steps of the hypothesis test

<br>

### Comparing two population proportions $\mu_1$ and $\mu_2$

$\bf (6.)$ A study published in the journal <i>Physical Culture and Sport. Studies and Research</i> explored what motivates athletes to keep playing sports, comparing team sports like football with individual ones like taekwondo [@moradi2020motivation]. The study used stratified random sampling to survey $265$ athletes from four team disciplines (football, volleyball, basketball, handball) and 2 individual disciplines (kung fu and taekwondo). The study evaluated motivational factors affecting sport participation based on eight fields: achievement/status, teamwork, fitness, energy release, situational factors, skill development, friendship, and fun. Each field was evaluated based on a survey that quantified sentiment in each field using three-level likert scale responses (very important=3, somewhat important=2, and not important=1). The results of the study are summarized in the table below: 

```{r, echo=F, warning=F, message=F}
tab = read.csv(file = paste0(datapath, 'sports_motivation_summary.csv'))
tab$Variables[tab$Variables == '\''] = ''

kable(tab, format = 'html', digits = 2, col.names = c('Motivational Factor', 'Sport Type', 'Sample Mean', 'Sample Stdev.', 'Degrees of Freedom'),
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')


```

Conduct a two independent samples $t$-test at the $\alpha = 0.05$ significance level to determine whether fitness as a motivating factor is significantly different between team and individual sports. Assume the populations have the SAME variance. 

<br>
<br>

$\bf (7.)$ A 1993 study published in the <i>Canadian Journal of Zoology</i> investigated the characteristics of maturation and growth in two lowland populations (Servotte and Thevenon) of European Common Frogs <i>Rana temporaria</i> [@augert1993plasticity] in Southeast France from 1986 - 1989. A summary of the body lengths for male and female frogs from 1 - 4 years in age in both populations are summarized below.


```{r, echo=F, warning=F, message=F}
tab = read.csv(file = paste0(datapath, 'sports_motivation_summary.csv'))
tab$Variables[tab$Variables == '\''] = ''
tab$DF = rep(c(203, 62), 9)

kable(tab, format = 'html', digits = 2, col.names = c('Motivational Factor', 'Sport Type', 'Sample Mean', 'Sample Stdev.', 'Sample Size'),
      row.names = F, booktabs = T, escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')


```


Conduct a two independent samples $t$-test at the $\alpha = 0.05$ significance level to evaluate whether there is a statistically significant difference in the mean body lengths of four-year-old male frogs between the Servotte and Thévenon populations. Assume the populations have DIFFERENT variances.

\newpage
# References
