---
title: "Homework 1 Solutions"
output:
  html_document: default
  pdf_document: default
always_allow_html: true
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=8, fig.pos = "H")
#knitr::opts_chunk$set(list(echo = FALSE, eval = FALSE))
library(kableExtra)
library(knitr)
library(latex2exp)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(plyr)
path = 'C:/Users/Bruin/Documents/GitHub/UI_STAT251-03_Spring_2024/docs/Data/'
set.seed(123)

#<span style="color: red;">text</span>
```




${\bf (1)}$	Define the following terms: 

* **(a)** Population - the collection of all possible persons, events, or objects of interest
* **(b)** Sample - the subset of the population that we actually observe (the observed observations) 
* **(c)** Parameter - A numerical characteristic of a population (alternatively: A function of the entire population) 
* **(d)** Statistic -  A numerical characteristic of a sample (alternatively: A function of the sample) 


${\bf (2)}$ Explain the difference between a **qualitative** (categorical) and a **quantitative** variable 

<span style="color: red;">A qualitative variable is a non-numeric characteristic such as a name or label while a quantitative variable is numerical characteristic such as a measurement or count</span>


${\bf (3)}$ Explain the difference between a **discrete** and a **continuous** variable and give an example of each

<span style="color: red;">A discrete variable is a quantitative variable that takes on only distinct whole number (integer) values such as counts whereas a continuous variable is a quantitative variable that can assume any value within a certain interval such temperature or height</span>


${\bf (4)}$ Explain the difference between a **nominal** and an **ordinal** variable and give an example of each

<span style="color: red;">A nominal variable is a qualitative variable that has categories with no inherent ordering such as the names of a city or the brands or shoes. An ordinal variable is a qualitative variable that has an inherent ordering to its categories such as a Likert score or education level.</span>

${\bf (5)}$ **At What age did women marry?**  A historian wants to estimate the average age at marriage of women in New England in the early 19th century. Within her state archives, she finds marriage records for the years $1800 - 1820$, which she treats as a sample of all marriage records from the early 19th century. The average age of the women in the records is $24.1$ years of age. Using the appropriate statistical method, she estimates that the average age of brides in the early $19^{th}$- century New England was between $23.5$ and $24.7$ years of age. 

* **(a)** Which part of this example gives a descriptive summary of the data?
<span style="color: red;">When she computes a mean of $24.1$ years of age from the marriage records</span>
* **(b)** Which part of this example draws an inference about a population?
<span style="color: red;">The part where she estimates the average age of women to be married in the 19th century to be between $23.5$ and $24.7$ years for the population</span>
* **(c)** What population is the historian is studying?
<span style="color: red;">All women who married in the 19th century}}
\item[(d)]{The average age the historian computed from the historical records was 24.1 years of age.</span>
* **(d)** The average age the historian computed from the historical records was 24.1 years of age. Is 24.1 years of age a statistic or a parameter? Why?
<span style="color: red;">A statistic. The quantity $24.1$ was computed from a sample of marriage records spanning the first $20$ years of the century</span>


${\bf (6)}$ Consider the following data obtained from flipping a coin 15 times. A value of $H$ denotes the coin was heads whereas a value of $T$ denotes the coin flip was tails.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
df = data.frame(`coin flip result:` = sample(c("H", "T"), 15, replace = T))

kable(t(df), format = 'html', booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
Compute the frequency table for the variable ``coin flip result" and answer the following questions:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
x1 = c('H', 'T')
x2 = c(7,8)
x3 = c('7/15 = 0.47','8/15 = 0.53')

kable(cbind.data.frame(x1, x2, x3), format = 'html', col.names = c('Result', 'Frequency', 'Relative Frequency'), booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
* **(a)** What type of variable is ``Coin Flip Result"?
<span style="color: red;">Qualitative nominal variable</span>
* **(b)** What proportion of the coin flips were heads?
<span style="color: red;">$0.47$ or $47\%$</span>
* **(c)** Why do we not compute the cumulative relative frequency for this variable?
<span style="color: red;">The variable is nominal and therefore has no inherent ordering for summing up the relative frequencies</span>
* **(d)** What the best graphical display for this variable and why?
<span style="color: red;">Since there are only two categories (heads or tails) a pie chart or bar graph would be sufficient to show the distribution</span>



${\bf (7)}$ A survey about color preferences reported the age distribution of the people who responded. Below are the results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

ag = c('1-18','19-24','25-25','36-60','51-69','70 and over')
cts = c(10, 97, 70, 36, 14, 5)
rf = c('<span style="color: red;">0.04</span>', 
       '<span style="color: red;">0.42</span>',
       '<span style="color: red;">0.30</span>',
       '<span style="color: red;">0.16</span>',
       '<span style="color: red;">0.06</span>',
       '<span style="color: red;">0.02</span>')

age.dist = cbind.data.frame(c('Age Group (Years)', 'Counts', '<span style="color: red;">Relative Frequency</span>'), rbind.data.frame(ag, cts, rf))

kable(age.dist, format = 'html', col.names = rep('', 7), booktabs = T, row.names = F,escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
Use this table to answer parts a - d

* **(a)** Compute the relative frequency for each age group

* **(b)** Make a bar graph where the heights of the bars are relative frequencies
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
set.seed(123)

ag = c('1-18','19-24','25-35','36-60','51-69','70 and over')
cts = c(10, 97, 70, 36, 14, 5)

rf = cts/sum(cts)

ggplot()+geom_bar(aes(x = ag, y = rf), stat = 'identity',
                  color = 'black', fill = 'lightblue')+
  theme_bw()+
  xlab('Age Group')+
  ylab('Relative Frequency')+
  theme(axis.text = element_text(size = 12))


```
* **(c)** Describe the distribution
<span style="color: red;">the distribution is slightly right skew indicating that most respondants were between 19 and 35 years of age</span>
* **(d)** Explain why your bar graph is not a histogram
<span style="color: red;">A histogram has intervals of equal length</span>

${\bf (8)}$ Email spam is the curse of the internet. The table below gives a compilation of the most common types of spam 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

`Type of Spam` = c('Adult','Financial','Health','Leisure','Products','Scams')
Percentage = c(14.5, 16.2, 7.3, 7.8, 21.0, 14.2)

spam.dist = cbind.data.frame(`Type of Spam`, Percentage)

kable(spam.dist, format = 'html', booktabs = T, row.names = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```

Use the table to answer the questions a and b

* **(a)** Report the modal spam category
<span style="color: red;">The modal category is "Products" because it has the highest frequency</span>
* **(b)** Construct a **Pareto chart** using the table above
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}

ggplot(data = spam.dist, aes(x = `Type of Spam`, y = sort(Percentage, decreasing = T)))+
  geom_bar(stat = 'identity', color = 'black', fill = 'lightblue2')+
  theme_hc()+
  xlab('Type of Spam')+
  ylab('Percentage')+
  ggtitle('Pareto Chart of Spam Categories')
```



${\bf (9)}$ A farmer in Idaho is interested in the number of rainy days in a given year so he records the number of rainy $R$, cloudy $C$ and sunny $S$ days over two weeks in $May$. His observations are $$\{R,R,C,C,C,S,S,C,R,R,C,R,S,S\}$$. Use the farmers observations to answer the following questions.  

* **(a)** Based on the farmers observations, what proportion of days were rainy? 
<span style="color: red;">$$\hat{p}_{\text{rainy}} = \frac{5}{14} = 0.36$$ </span>
* **(b)** Is the proportion in part **(a)** a statistic or a parameter? why?
<span style="color: red;">It is a statistic because it was computed from a sample of 14 observations of the weather</span>
* **(c)** What population is the farmer studying? is it a finite or infinite population?
<span style="color: red;">The population is all 365 days in a given year. It is finite</span>


${\bf (10)}$ Consider the following data from a survey conducted on college students in the state of Florida. As part of their research, surveyors recorded the high school performance (measured in grade point average - GPA) of 35 college students from across the state. For your convenience the data have been sorted from least to greatest:

$$2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,$$ 
$$3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,$$ 
$$3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,$$ 
$$3.8, 3.8, 3.8, 4.0, 4.0$$


The following frequency table gives the distribution of the variable high school GPA (with values rounded to nearest 0.5). Fill in the table and answer the following questions:

```{r, fig.cap="Frequency Table for GPA after rounding values to the nearest 0.5",  echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
#fldata = read.csv(paste0(path, 'fl_student_survey.csv'))
#fldata$gpa_rounded = round_any(fldata$high_sch_GPA, 0.5)
x_raw = c(2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,3.8, 3.8, 3.8, 4.0, 4.0)
x=round_any(x_raw, 0.5)
hsGPA_values = sort(unique(x))
hsGPA_freq = summary(as.factor(x))
hsGPA_RF = hsGPA_freq/sum(hsGPA_freq)
hsGPA_cRF = cumsum(hsGPA_RF)


FT = cbind.data.frame(GPA = hsGPA_values, 
                      Frequency = hsGPA_freq,
                      `Relative frequency` = round(hsGPA_RF,2),
                      `Cumulative RF` = round(hsGPA_cRF, 2))

kable(FT, format = 'html', digits = 2, row.names = F, booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')
                      

```


* **(a)** What type of variable is GPA?
<span style="color: red;">Quantitative continuous</span>
* **(b)** What proportion of college students had a high school GPA $> 3.0$?
<span style="color: red;">0.57</span>
* **(c)** What proportion of college students had a high school GPA $< 3.0$?
<span style="color: red;">0.09</span>
* **(d)** What is the **mean** GPA in this sample?
<span style="color: red;">$$\bar{x} = \frac{2.0+2.1+...+4.0+4.0}{35} = \frac{2(2)+1(2.5)+...+6(4)}{35} = 3.3$$</span>
* **(e)** What is the **median** GPA in this sample?
<span style="color: red;">median $= 3.4$</span>
* **(f)** What is the **mode** of GPA in this sample?

<span style="color: red;">Using the frequency table on the rounded values the mode is the value with the highest relative frequency $ = 3.5$</span>

<span style="color: red;">(Using the raw data) the mode is the most frequent value observed in the data which is $3.0$ </span>


* **(g)** Construct a dot plot for the variable GPA (hint use the frequency table)
```{r, fig.cap = "(A) Dot plot using the rounded GPA values from the frequency table (B) Dot plot using the raw GPA values", fig.height=3, fig.width=7, echo=FALSE, message = FALSE, warning=FALSE}

df = cbind.data.frame(Raw = x_raw, Rounded = x)
B = ggplot(data = df, aes(x = Rounded))+
  geom_dotplot(dotsize = 1)+
  theme_classic()+
  ylab("")+
  xlab("Rounded Student High School GPA")+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank())
A = ggplot(data = df, aes(x = Raw))+
  geom_dotplot(dotsize = 1)+
  theme_classic()+
  ylab("")+
  xlab("Student High School GPA")+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank())


ggarrange(B, A, labels = c("A", "B"), nrow = 1, ncol = 2)
  
```

* **(h)** Construct a histogram for the variable GPA (hint use the frequency table)
```{r, fig.cap = "(A) Histogram using the rounded GPA values (e.g 4 total bins) from the frequency table", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE, comment="", results="asis"}
A = ggplot(data = df, aes(x = Rounded))+
  geom_histogram(breaks = seq(2, 4, 0.5), color = 'black', fill = 'grey')+
  theme_classic()+
  ylab("Frequency")+
  xlab("Rounded Student High School GPA")+
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14))
plot(A)
```

* **(i)** Construct a stem and leaf plot for the variable GPA
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}
print("Using the Rounded GPA scores:")
print(stem(df$Rounded))
print("Using the Raw Data:")
print(stem(df$Raw, scale = 3))
```

${\bf (11)}$ Which statistic is more resistant to outliers, the mean or median? Why?

<span style="color: red;">The median is resistant to outliers because it relies on only the middle value (or middle two values if $n$ is even). Therefore, how far a value is from the center does not influence the median. The mean, however, is computed using \textbf{ALL} values in a sample distribution. This causes the mean to be pulled in the direction of extreme values.</span>

${\bf (12)}$ Describe the shape of the following distributions and for each distribution identify if the mean will be larger, smaller or the same as the median.

**(a)**
<span style="color: red;">Skew right. The mean will be greater than median because the center of the data is pulled toward the right side of the distribution</span>


```{r, fig.height=3, fig.width=4, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
sr = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 2, 10)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sr)
```


**(b)**   
<span style="color: red;">Symmetric/Bell shaped. The mean and the median will be equal</span>


```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sym = ggplot()+
  geom_histogram(aes(x = rnorm(100000)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sym)
```


**(c)**   
<span style="color: red;">Skew left. The mean will be less than median because the center of the data is pulled toward the left side of the distribution</span>


```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sl = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 10, 2)), fill = 'lightblue', size = 1)+
  theme_base()+
  # geom_vline(xintercept = mean(rbeta(100000, 10, 2)), linetype = 'dotted', color = 'red')+
  # geom_text(aes(x = mean(rbeta(100000, 10, 2)), y = 0), label = 'mean', color = 'red')+
  # geom_vline(xintercept = median(rbeta(100000, 10, 2)), linetype = 'dotted', color = 'green')+
  # geom_text(aes(x = mean(rbeta(100000, 10, 2)), y = 0), label = 'median', color = 'green')+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sl)
```



\newpage
${\bf (13)}$ Consider the following four sets of observations of a quantitative variable $x$. For your convenience the observations have been sorted in increasing order. Match datasets $1-4$ with the correct histogram (labeled $A - D$)

$$\text{Dataset 1} = \{0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6\}$$
$$\text{Dataset 2}= \{0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3\}$$
$$\text{Dataset 3} = \{1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7\}$$
$$\text{Dataset 4} = \{3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7\}$$
```{r, echo=FALSE, message=FALSE, warning=F}

set.seed(123)


a = c(0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6)
b = c(0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3)
c = c(1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7)
d = c(3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7)

par(mfrow = c(2,2))
hist(a, xlab = "", breaks = seq(0, 10, 2), main = "Distribution A")
hist(c, xlab = "", breaks = seq(0, 10, 2), main = "Distribution B")
hist(d, xlab = "", breaks = seq(0, 10, 2), main = "Distribution C")
hist(b, xlab = "", breaks = seq(0, 10, 2), main = "Distribution D")
par(mfrow = c(1,1))
```


* <span style="color: red;">Dataset 1 goes with histogram A</span> 
* <span style="color: red;">Dataset 2 goes with histogram D</span>
* <span style="color: red;">Dataset 3 goes with histogram B</span>
* <span style="color: red;">Dataset 4 goes with histogram C</span>


${\bf (14)}$ Consider the following set of 10 observations of a variable $X$ sorted from least to greatest:
$$3.3, 3.8, 4.0, 4.8, 4.8, 5.1, 5.2, 5.6, 5.7, 6.9$$
Use the data to answer parts a-b

* **(a)** Given the data, compute the $50^{th}$ percentile of $X$ 
<span style="color: red;">$$\frac{4.8+5.1}{2} = 4.95$$</span>
* **(b)** Compute Interquartile Range (IQR) range of $X$
<span style="color: red;">$$IQR = 5.6 - 4.0 = 1.6$$</span>


${\bf (15)}$ Consider the following $n = 20$ observations of the sugar and sodium content of several popular cereal brands and answer questions a - g:

```{r, echo = FALSE, message=FALSE, warning=FALSE}
cereal = read.csv(paste0(path, 'cereal.csv'))
kable(cereal, format = 'html', col.names = c("Brand", "Sodium (mg)", "Sugar (g)", 'Type'))%>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, width = "18em")%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```

* **(a)** Construct a histogram of the data using the following bins: $\leq 2 (g)$, $2 - 4 (g)$, 
$4 - 6 (g)$, $6 - 8 (g)$,$8 - 10 (g)$, $10 - 12 (g)$, $12 - 14 (g)$,$14 - 16 (g)$,$> 16 (g)$ (hint start by creating a frequency table) 

```{r, fig.cap = "(A) Histogram of the variable sugar (g) using the bins in the frequency table (above)", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE}
A = ggplot(data = cereal, aes(x = Sugar))+
  geom_histogram(breaks = seq(0, 20, 2), color = 'black', fill = 'grey')+
  theme_classic()+
  ylab("Frequency")+
  xlab("Student High School GPA")+
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14))
plot(A)

x_rounded = hist(cereal$Sugar, breaks = seq(0, 20, 2), plot = F)
x_rf = x_rounded$counts/sum(x_rounded$counts)
x_crf = cumsum(x_rf)

Tabular = cbind.data.frame(`Sugar (g)` = c("< = 2 (g)", 
                                           "2 - 4 (g)",
                                           "4 - 6 (g)",
                                           "6 - 8 (g)",
                                           "8 - 10 (g)",
                                           "10 - 12 (g)",
                                           "12 - 14 (g)",
                                           "14 - 16 (g)",
                                           "> 16 (g)"),
                           Frequency = x_rounded$counts[1:9],
                           `RF(x)` = x_rf[1:9],
                           `CRF(x)` = x_crf[1:9])

kable(Tabular, caption = "Frequency table used to construct histogram for the variable \"Sugar (g)\"")
  
```


<br>

* **(b)** Compute the mean, median, and mode of the variable ``Sugar (g)"
<span style="color: red;">Using the raw data there are 5 modes each with a frequency of 2
mode(s) $= 3,4,11,12,14$</span>
<span style="color: red;">median $= \frac{9+10}{2} = 9.5$</span>
<span style="color: red;">$$\bar{x} = \frac{1}{20}\sum_{i=1}^{20} x_i = \frac{(11+18+5+14+...+3+11)}{20} = 8.75$$ (note that the mean may be slightly different if you used the frequency table above</span>

* **(c)** Compute the variance and standard deviation of the variable ``Sugar (g)"
<span style="color: red;">$$s^2 = \frac{1}{19}\sum_{i=1}^{19} (x-\bar{x})^2 = \frac{(11 - 8.75)^2 + (18-8.75)^2 + ...+(3-8.75)^2 + (11 - 8.75)^2}{19}$$  $$ = \frac{5.06 + 85.56 + ...+33.06+5.06}{19} = 28.3$$ $$s = \sqrt{s^2} = \sqrt{28.3} = 5.32$$</span>


* **(d)** Compute the quartiles (Q1, Q2, Q3) and interquartile range of the variable ``Sugar (g)"
<span style="color: red;">$$Q1 = 4, \ Q2 = \text{median} = 9.5, \ Q3 = 13, \ IQR = 13 - 4 = 9$$</span>

* **(e)** Create a box and whisker plot for the variable ``Sugar (g)" and mark the Q1, Q2, and Q3 quartiles, the median and any potential outliers

```{r, fig.cap = "Boxplot of Sugar content in 20 cereal brands", fig.height=5, fig.width=6, echo=FALSE, message = FALSE, warning=FALSE}
X = cereal$Sugar

mi = min(X)
mx = max(X)
Q1 = (4+4)/2
med = 9.5
Q3 = (12+14)/2
estims = c(mi, Q1, med, Q3, mx)
boxplot(X, horizontal = TRUE, axes = TRUE, staplewex = 1)
text(x=estims-0.1, labels =paste(c("Min", "Q1","Median", "Q3", "Max"), 
                                   estims, sep = "="), y=1.25, cex = 0.68)
text(x=Q3-Q1, labels =paste("IQR", Q3 - Q1, sep = " = "), y=1.5, cex = 0.68)
```

* **(f)** Plot the Cumulative Distribution of the variable ``Sugar (g)"

```{r, fig.cap = "Cumulative Distribution of Sugar (g)", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE}

A=ggplot()+geom_step(aes(x = x_rounded$breaks[2:10], y = x_crf[1:9]), size = 0.8)+
  geom_point(aes(x = x_rounded$breaks[2:10], y = x_crf[1:9]), size = 3)+
  theme_hc()+
  xlab('Sugar (g)')+
  ylab('Cumulative Relative Frequency')+
  scale_x_continuous(breaks = seq(0, 20, 2))+
  scale_y_continuous(breaks = seq(0, 1, 0.1))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))
  
  plot(A)
```

* **(g)** What proportion of cereals have a sugar content less than or equal to 10 grams?
<span style="color: red;">$$55\%$$</span>



<!-- **(16)** True or False: The range is the best descriptive statistic for the spread of a distribution (Explain your answer) -->

<!-- **(17)** True or False: The standard deviation $s$ is the square of the variance -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->

<!-- **(18)** True or False: A **deviation} is defined as the most frequent value in a distribution -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->













