---
title: "Homework 1 Problems"
output:
  html_document: default
  pdf_document: default
always_allow_html: true
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=8, fig.pos = "H")
#knitr::opts_chunk$set(list(echo = FALSE, eval = FALSE))
library(kableExtra)
library(knitr)
library(latex2exp)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(plyr)
path = 'C:/Users/Bruin/Documents/GitHub/UI_STAT251-03_Spring_2024/docs/Data/'
set.seed(123)
```




${\bf (1)}$	Define the following terms:
* **(a)** Population
* **(b)** Sample
* **(c)** Parameter
* **(d)** Statistic


${\bf (2)}$ Explain the difference between a **qualitative** (categorical) and a **quantitative** variable


${\bf (3)}$ Explain the difference between a **discrete** and a **continuous** variable and give an example of each


${\bf (4)}$ Explain the difference between a **nominal** and an **ordinal** variable and give an example of each

${\bf (5)}$ **At What age did women marry?**  A historian wants to estimate the average age at marriage of women in New England in the early 19th century. Within her state archives, she finds marriage records for the years $1800 - 1820$, which she treats as a sample of all marriage records from the early 19th century. The average age of the women in the records is $24.1$ years of age. Using the appropriate statistical method, she estimates that the average age of brides in the early $19^{th}$- century New England was between $23.5$ and $24.7$ years of age. 

* **(a)** Which part of this example gives a descriptive summary of the data?
* **(b)** Which part of this example draws an inference about a population?
* **(c)** What population is the historian is studying?
* **(d)** The average age the historian computed from the historical records was 24.1 years of age. Is 24.1 years of age a statistic or a parameter? Why?


${\bf (6)}$ Consider the following data obtained from flipping a coin 15 times. A value of $H$ denotes the coin was heads whereas a value of $T$ denotes the coin flip was tails.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
df = data.frame(`coin flip result:` = sample(c("H", "T"), 15, replace = T))

kable(t(df), format = 'html', booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
Compute the frequency table for the variable ``coin flip result" and answer the following questions:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
x1 = c('H', 'T')
x2 = c('','')
x3 = c('','')

kable(cbind.data.frame(x1, x2, x3), format = 'html', col.names = c('Result', 'Frequency', 'Relative Frequency'), booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
* **(a)** What type of variable is ``Coin Flip Result"?
* **(b)** What proportion of the coin flips were heads?
* **(c)** Why do we not compute the cumulative relative frequency for this variable?
* **(d)** What the best graphical display for this variable and why?



${\bf (7)}$ A survey about color preferences reported the age distribution of the people who responded. Below are the results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

ag = c('1-18','19-24','25-35','36-60','51-69','70 and over')
cts = c(10, 97, 70, 36, 14, 5)

age.dist = cbind.data.frame(c('Age Group (Years)', 'Counts'), rbind.data.frame(ag, cts))

kable(age.dist, format = 'html', col.names = rep('', 7), booktabs = T, row.names = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```
Use this table to answer parts a - d

* **(a)** Compute the relative frequency for each age group
* **(b)** Make a bar graph where the heights of the bars are relative frequencies
* **(c)** Describe the distribution
* **(d)** Explain why your bar graph is not a histogram

${\bf (8)}$ Email spam is the curse of the internet. The table below gives a compilation of the most common types of spam 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

`Type of Spam` = c('Adult','Financial','Health','Leisure','Products','Scams')
Percentage = c(14.5, 16.2, 7.3, 7.8, 21.0, 14.2)

spam.dist = cbind.data.frame(`Type of Spam`, Percentage)

kable(spam.dist, format = 'html', booktabs = T, row.names = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```

Use the table to answer the questions a and b

* **(a)** Report the modal spam category
* **(b)** Construct a **Pareto chart** using the table above





${\bf (9)}$ A farmer in Idaho is interested in the number of rainy days in a given year so he records the number of rainy $R$, cloudy $C$ and sunny $S$ days over two weeks in $May$. His observations are $$\{R,R,C,C,C,S,S,C,R,R,C,R,S,S\}$$. Use the farmers observations to answer the following questions.  

* **(a)** Based on the farmers observations, what proportion of days were rainy? 

* **(b)** Is the proportion in part **(a)** a statistic or a parameter? why?

* **(c)** What population is the farmer studying? is it a finite or infinite population?



${\bf (10)}$ Consider the following data from a survey conducted on college students in the state of Florida. As part of their research, surveyors recorded the high school performance (measured in grade point average - GPA) of 35 college students from across the state. For your convenience the data have been sorted from least to greatest:

$$2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,$$ 
$$3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,$$ 
$$3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,$$ 
$$3.8, 3.8, 3.8, 4.0, 4.0$$


The following frequency table gives the distribution of the variable high school GPA (with values rounded to nearest 0.5). Fill in the table and answer the following questions:

```{r, fig.cap="Frequency Table for GPA after rounding values to the nearest 0.5",  echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
#fldata = read.csv(paste0(path, 'fl_student_survey.csv'))
#fldata$gpa_rounded = round_any(fldata$high_sch_GPA, 0.5)

x=round_any(c(2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,3.8, 3.8, 3.8, 4.0, 4.0), 0.5)
hsGPA_values = sort(unique(x))
hsGPA_freq = summary(as.factor(x))
hsGPA_RF = hsGPA_freq/sum(hsGPA_freq)
hsGPA_cRF = cumsum(hsGPA_RF)


FT = cbind.data.frame(GPA = hsGPA_values, 
                      Frequency = c(hsGPA_freq[1:2], "", hsGPA_freq[4:5]),
                      `Relative frequency` = c(round(hsGPA_RF[1],2), "", "", round(hsGPA_RF[4:5],2)),
                      `Cumulative RF` = c("", "", round(hsGPA_cRF[3], 2), "", ""))

kable(FT, format = 'html', digits = 2, row.names = F, booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')
                      

```


* **(a)** What type of variable is GPA?
* **(b)** What proportion of college students had a high school GPA $> 3.0$?
* **(c)** What proportion of college students had a high school GPA $< 3.0$?
* **(d)** What is the **mean** GPA in this sample?
* **(e)** What is the **median** GPA in this sample?
* **(f)** What is the **mode** of GPA in this sample?
* **(g)** Construct a dot plot for the variable GPA (hint use the frequency table)
* **(h)** Construct a histogram for the variable GPA (hint use the frequency table)
* **(i)** Construct a stem and leaf plot for the variable GPA



${\bf (11)}$ Which statistic is more resistant to outliers, the mean or median? Why?


${\bf (12)}$ Describe the shape of the following distributions and for each distribution identify if the mean will be larger, smaller or the same as the median.

**(a)**

```{r, fig.height=3, fig.width=4, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
sr = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 2, 10)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sr)
```

**(b)**   

```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sym = ggplot()+
  geom_histogram(aes(x = rnorm(100000)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sym)
```

**(c)**   

```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sl = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 10, 2)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sl)
```



\newpage
${\bf (13)}$ Consider the following four sets of observations of a quantitative variable $x$. For your convenience the observations have been sorted in increasing order. Match datasets $1-4$ with the correct histogram (labeled $A - D$)

$$\text{Dataset 1} = \{0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6\}$$
$$\text{Dataset 2}= \{0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3\}$$
$$\text{Dataset 3} = \{1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7\}$$
$$\text{Dataset 4} = \{3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7\}$$
```{r, echo=FALSE, message=FALSE, warning=F}

set.seed(123)


a = c(0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6)
b = c(0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3)
c = c(1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7)
d = c(3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7)

par(mfrow = c(2,2))
hist(a, xlab = "", breaks = seq(0, 10, 2), main = "Distribution A")
hist(c, xlab = "", breaks = seq(0, 10, 2), main = "Distribution B")
hist(d, xlab = "", breaks = seq(0, 10, 2), main = "Distribution C")
hist(b, xlab = "", breaks = seq(0, 10, 2), main = "Distribution D")
par(mfrow = c(1,1))
```



${\bf (14)}$ Consider the following set of 10 observations of a variable $X$ sorted from least to greatest:
$$3.3, 3.8, 4.0, 4.8, 4.8, 5.1, 5.2, 5.6, 5.7, 6.9$$
Use the data to answer parts a-b

* **(a)** Given the data, compute the $40^{th}$ percentile of $X$ 
* **(b)** Compute Interquartile Range (IQR) range of $X$


${\bf (15)}$ Consider the following $n = 20$ observations of the sugar and sodium content of several popular cereal brands and answer questions a - g:

```{r, echo = FALSE, message=FALSE, warning=FALSE}
cereal = read.csv(paste0(path, 'cereal.csv'))
kable(cereal, format = 'html', col.names = c("Brand", "Sodium (mg)", "Sugar (g)", 'Type'))%>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, width = "18em")%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped')

```

* **(a)** Construct a histogram of the data using the following bins: $\leq 2 (g)$, $2 - 4 (g)$, 
$4 - 6 (g)$, $6 - 8 (g)$,$8 - 10 (g)$, $10 - 12 (g)$, $12 - 14 (g)$,$14 - 16 (g)$,$> 16 (g)$ (hint start by creating a frequency table) 
* **(b)** Compute the mean, median, and mode of the variable ``Sugar (g)"
* **(c)** Compute the variance and standard deviation of the variable ``Sugar (g)"
* **(d)** Compute the quartiles (Q1, Q2, Q3) and interquartile range of the variable ``Sugar (g)"
* **(e)** Create a box and whisker plot for the variable ``Sugar (g)" and mark the Q1, Q2, and Q3 quartiles, the median and any potential outliers
* **(f)** Plot the Cumulative Distribution of the variable ``Sugar (g)"
* **(g)** What proportion of cereals have a sugar content less than or equal to 10 grams?



<!-- **(16)** True or False: The range is the best descriptive statistic for the spread of a distribution (Explain your answer) -->

<!-- **(17)** True or False: The standard deviation $s$ is the square of the variance -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->

<!-- **(18)** True or False: A **deviation} is defined as the most frequent value in a distribution -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->













